A Framework for Substructural Type Systems \thanks{James Wood is supported by an EPSRC Studentship.}

### 2022-01-24 agda/LRKits.lagda.tex
[ edit ] minor updates

  open import Generic.Linear.Operations rawPoSemiring
  open import Generic.Linear.Algebra poSemiring
  open import Generic.Linear.Syntax Ty Ann
  open import Generic.Linear.Syntax.Interpretation Ty rawPoSemiring
  open import Generic.Linear.Variable Ty rawPoSemiring
  open import Generic.Linear.Environment Ty poSemiring
  open import Generic.Linear.Environment.Properties Ty poSemiring
  open import Generic.Linear.Environment.Categorical Ty poSemiring
\end{code}

@@

    env-+ : ∀ {s t γ δ P Qₗ Qᵣ} →
      let Γ = ctx {s} P γ; Δₗ = ctx {t} Qₗ δ; Δᵣ = ctx {t} Qᵣ δ in
      (λ Q → [ 𝓥 ] Γ ⇒ᵉ ctx Q δ) ◇ _≤[ Qₗ +* Qᵣ ] →
      (λ Pₗ → [ 𝓥 ] ctx Pₗ γ ⇒ᵉ Δₗ)
        ↘- P ≤[_+*_] -↙
      (λ Pᵣ → [ 𝓥 ] ctx Pᵣ γ ⇒ᵉ Δᵣ)
    env-+ (ρ , sp) =
      let sumsₗ ↘, sp′ ,↙ sumsᵣ = ρ .asLinRel .linRel .rel-+ₘ (sp , ρ .sums) in
      record { [_]_⇒ᵉ_ ρ; sums = sumsₗ }
        ↘, sp′ ,↙
      record { [_]_⇒ᵉ_ ρ; sums = sumsᵣ }

    env-* : ∀ {s t γ δ P Q′ r} →
      let Γ = ctx {s} P γ; Δ′ = ctx {t} Q′ δ in
      (λ Q → [ 𝓥 ] Γ ⇒ᵉ ctx Q δ) ◇ _≤[ r *ₗ Q′ ] →
      P ≤[ r *ₗ_] ◇ (λ P′ → [ 𝓥 ] ctx P′ γ ⇒ᵉ Δ′)
    env-* (ρ , sp) =
      let sp′ , sums′ = ρ .asLinRel .linRel .rel-*ₘ (sp , ρ .sums) in
      sp′ , record { [_]_⇒ᵉ_ ρ; sums = sums′ }
\end{code}

@@

\begin{code}
    open Equivalence
\end{code}

%<*bindEnv>
\begin{code}
    bindEnv : ∀ {Γ Δ Θ} → [ 𝓥 ] Γ ⇒ᵉ Δ → [ 𝓥 ] Γ ++ᶜ Θ ⇒ᵉ Δ ++ᶜ Θ
    bindEnv ρ .M = [ [ ρ .M │ 0ᴹ ] ─ [ 0ᴹ │ 1ᴹ ] ]
    bindEnv ρ .asLinRel = ρ .asLinRel ⊕AsLinRel idAsLinRel
    bindEnv ρ .sums = ρ .sums , ≤*-refl
    bindEnv ρ .lookup (ll , rr) (lvar (↙ i) q b) =
      let bl , br = un++ₙ b in
      ↙ᵏ (≤*-trans rr br) (ρ .lookup ll (lvar i q bl))
    bindEnv ρ .lookup (ll , rr) (lvar (↘ i) q b) =
      let bl , br = un++ₙ b in
      let bl′ = 0*→≤* (ρ .asLinRel .linRel .rel-0ₘ (≤*→0* bl , ll)) in
      vr (lvar (↘ i) q (bl′ ++ₙ ≤*-trans rr br))
\end{code}
%</bindEnv>

%<*trav>
\begin{code}
    trav : ∀ {Γ Δ A} → [ 𝓥 ] Γ ⇒ᵉ Δ → Δ ⊢ A → Γ ⊢ A
    trav ρ (var x) = tm (ρ .lookup (ρ .sums) x)
    trav ρ (!-I (⟨ sp* ⟩· M)) =
      let sp*′ , ρ′ = env-* (ρ , sp*) in
      !-I (⟨ sp*′ ⟩· (trav ρ′ M))
    trav ρ (!-E (M ✴⟨ sp+ ⟩ N)) =
      let ρₗ ↘, sp+′ ,↙ ρᵣ = env-+ (ρ , sp+) in
      !-E (trav ρₗ M ✴⟨ sp+′ ⟩ trav (bindEnv ρᵣ) N)
\end{code}
%</trav>

@@

%<*dlv-env>
\begin{code}
  ↙^Env : {𝓥 : Ctx → Ty → Set} → Weakening 𝓥 → Weakening [ 𝓥 ]_⇒ᵉ_
  ↙^Env ↙^𝓥 R0 ρ .M = [ ρ .M │ 0ᴹ ]
  ↙^Env ↙^𝓥 R0 ρ .asLinRel = [ ρ .asLinRel │ 0AsLinRel ]AsLinRel
  ↙^Env ↙^𝓥 R0 ρ .sums = ρ .sums , ≤*→0* R0
  ↙^Env ↙^𝓥 R0 ρ .lookup (r , sp0) v = ↙^𝓥 (0*→≤* sp0) (ρ .lookup r v)

>>>>

  open import Generic.Linear.Operations rawPoSemiring
  open import Generic.Linear.Algebra poSemiring
  open import Generic.Linear.Syntax Ty Ann
  open import Generic.Linear.Syntax.Interpretation Ty rawPoSemiring
  open import Generic.Linear.Variable Ty rawPoSemiring hiding (_↙ᵛ_; _↘ᵛ_)
  open import Generic.Linear.Environment Ty poSemiring
  open import Generic.Linear.Environment.Properties Ty poSemiring
  open import Generic.Linear.Environment.Categorical Ty poSemiring
\end{code}

@@

    env-+ : ∀ {s t γ δ P Qₗ Qᵣ} →
      let Γ = ctx {s} P γ; Δₗ = ctx {t} Qₗ δ; Δᵣ = ctx {t} Qᵣ δ in
      (λ Q → [ 𝓥 ] Γ ⇒ᵉ ctx Q δ) ◇ _≤[ Qₗ +* Qᵣ ] →
      (λ Pₗ → [ 𝓥 ] ctx Pₗ γ ⇒ᵉ Δₗ)
        ↘- P ≤[_+*_] -↙
      (λ Pᵣ → [ 𝓥 ] ctx Pᵣ γ ⇒ᵉ Δᵣ)
    env-+ (ρ , sp) =
      let fit-hereₗ ↘, sp′ ,↙ fit-hereᵣ = ρ .Ψ .rel-+ₘ (sp , ρ .fit-here) in
      record { [_]_⇒ᵉ_ ρ; fit-here = fit-hereₗ }
        ↘, sp′ ,↙
      record { [_]_⇒ᵉ_ ρ; fit-here = fit-hereᵣ }

    env-* : ∀ {s t γ δ P Q′ r} →
      let Γ = ctx {s} P γ; Δ′ = ctx {t} Q′ δ in
      (λ Q → [ 𝓥 ] Γ ⇒ᵉ ctx Q δ) ◇ _≤[ r *ₗ Q′ ] →
      P ≤[ r *ₗ_] ◇ (λ P′ → [ 𝓥 ] ctx P′ γ ⇒ᵉ Δ′)
    env-* (ρ , sp) =
      let fit-here′ , sp′ = ρ .Ψ .rel-*ₘ (sp , ρ .fit-here) in
      sp′ , record { [_]_⇒ᵉ_ ρ; fit-here = fit-here′ }
\end{code}

@@

\begin{code}
    open Equivalence
    private module Vᶠᴿ {s} = FRelLeftSemimodule (Vᶠᴿ s)
    open Vᶠᴿ
\end{code}

%<*bindEnv>
\begin{code}
    bindEnv : ∀ {Γ Δ Θ} → [ 𝓥 ] Γ ⇒ᵉ Δ → [ 𝓥 ] Γ ++ᶜ Θ ⇒ᵉ Δ ++ᶜ Θ
    bindEnv ρ .Ψ = [ [ ρ .Ψ │ 0ᴿ ]ᴿ ─ [ 0ᴿ │ 1ᴿ ]ᴿ ]ᴿ
    bindEnv ρ .fit-here = _↘,_,↙_
      {left = _ ++ _} {right = _ ++ _}
      (ρ .fit-here , 0*-triv)
      (+*-identity↘ _ ++ₙ +*-identity↙ _)
      (0*-triv , ≤*-refl)
    bindEnv ρ .lookup ((ul , ur) ↘, sp+ ,↙ (dl , dr)) (lvar i q b)
      with un++ₙ sp+ | un++ₙ b
    bindEnv ρ .lookup ((ul , ur) ↘, sp+ ,↙ (dl , dr)) (lvar (↙ i) q b)
      | sp+l , sp+r | bl , br = ↙ᵏ
      (≤*-trans (+ₘ-identityˡ→ (ur , sp+r)) (≤*-trans dr br))
      (ρ .lookup (ρ .Ψ .rel-≤ₘ ≤*-refl (+ₘ-identityʳ→ (sp+l , dl)) ul)
        (lvar i q bl))
    bindEnv ρ .lookup ((ul , ur) ↘, sp+ ,↙ (dl , dr)) (lvar (↘ i) q b)
      | sp+l , sp+r | bl , br = vr (lvar (↘ i) q
      (≤*-trans (+ₘ-identityʳ→ (sp+l , dl))
        (0*→≤* (ρ .Ψ .rel-0ₘ (≤*→0* bl , ul)))
      ++ₙ
      ≤*-trans (+ₘ-identityˡ→ (ur , sp+r)) (≤*-trans dr br)))
\end{code}
%</bindEnv>

%<*trav>
\begin{code}
    trav : ∀ {Γ Δ A} → [ 𝓥 ] Γ ⇒ᵉ Δ → Δ ⊢ A → Γ ⊢ A
    trav ρ (var x) = tm (ρ .lookup (ρ .fit-here) x)
    trav ρ (!-I (⟨ sp* ⟩· M)) =
      let sp*′ , ρ′ = env-* (ρ , sp*) in
      !-I (⟨ sp*′ ⟩· (trav ρ′ M))
    trav ρ (!-E (M ✴⟨ sp+ ⟩ N)) =
      let ρₗ ↘, sp+′ ,↙ ρᵣ = env-+ (ρ , sp+) in
      !-E (trav ρₗ M ✴⟨ sp+′ ⟩ trav (bindEnv ρᵣ) N)
\end{code}
%</trav>

@@

%<*dlv-env>
\begin{code}
  ↙^Env : {𝓥 : Ctx → Ty → Set} → Weakening 𝓥 → Weakening [ 𝓥 ]_⇒ᵉ_
  ↙^Env ↙^𝓥 R0 ρ .Ψ = [ ρ .Ψ │ 0ᴿ ]ᴿ
  ↙^Env ↙^𝓥 R0 ρ .fit-here = ρ .fit-here , ≤*→0* R0
  ↙^Env ↙^𝓥 R0 ρ .lookup (r , sp0) v = ↙^𝓥 (0*→≤* sp0) (ρ .lookup r v)

### 2022-01-24 tex/esop22/algebra.tex
[ edit ] minor updates

The basic algebraic structure we concern ourselves with is \emph{partially
ordered semirings}, or \emph{posemirings} for short.
A posemiring is a (not necessarily commutative) semiring on a partially ordered
set, where both operations are monotonic.
As in many similar formalisms, posemiring elements represent usage restrictions,
with addition collecting restrictions from multiple uses, multiplication
handling usage under a modality, and the order giving subsumption of
restrictions, comparable to subtyping.

\begin{definition}
  A \emph{posemiring} is a tuple $(\Ann, \leq, 0, +, 1, *)$ such that
  $(\Ann, \leq)$ is a partially ordered set, $(\Ann, 0, +)$ is a commutative
  monoid, $(\Ann, 1, *)$ is a monoid, $+$ and $*$ are monotonic, and $*$
  distributes over $0$ and $+$ in the usual way on both sides.
\end{definition}

@@

Vectors of a given length form a \emph{module} over the posemiring $\Ann$,
analogously to how vectors over a field form a vector space.

>>>>

The basic algebraic structure we deal with is \emph{partially
ordered semirings}, or \emph{posemirings} for short.
A posemiring is a (not necessarily commutative) semiring on a partially ordered
set, where both operations are monotonic.
As in many similar formalisms, posemiring elements represent usage restrictions,
with addition collecting restrictions from multiple uses, multiplication
handling usage under a modality, and the order giving subsumption of
restrictions, comparable to subtyping.

\begin{definition}
  A \emph{posemiring} is a tuple $(\Ann, \leq, 0, +, 1, *)$ such that
  $(\Ann, \leq)$ is a partially ordered set, $(\Ann, 0, +)$ is a commutative
  monoid, $(\Ann, 1, *)$ is a monoid, $+$ and $*$ are monotonic, and $*$
  distributes over $0$ and $+$ on both sides.
\end{definition}

@@

Vectors of a given length form a \emph{module} over the posemiring $\Ann$,
analogously to how vectors over a field form a vector space.
The partial order on such vectors is pointwise.

### 2022-01-24 tex/esop22/examples.tex
[ edit ] minor updates

In an unpublished note, \citet{McBride05} gives a parametrised traversal
yielding homomorphisms of syntax.
The parameters are collected in the record \AgdaRecord{Kit}.
We make a minor change to the original presentation, where instead of our
\AgdaField{ren\textasciicircum{}$\V$} field, \citeauthor{McBride05} has the
field \AgdaField{wk} allowing only context extensions.
As for the other two fields, \AgdaField{vr} allows us to map variables to
$\V$-values, so as to put newly bound variables in environments; and
\AgdaField{tm} allows us to extract terms from $\V$-values, as required when
we use the environment to evaluate a free variable.

@@

The idea of renaming is that variables replace variables, whereas with
substitution, terms replace variables.
This translates to environments for renaming containing $\sqni$-values
(variables), and environments for substitution containing $\vdash$-values
(terms).

>>>>

In an unpublished note, \citet{McBride05} gives a parametrised traversal
yielding homomorphisms of syntax, the canonical examples of which are
simultaneous renaming and simultaneous substitution.
The parameters are collected in the record \AgdaRecord{Kit}.
We make a minor change to the original presentation, where instead of our
\AgdaField{ren\textasciicircum{}$\V$} field, \citeauthor{McBride05} has the
field \AgdaField{wk} allowing only context extensions.
As for the other two fields, \AgdaField{vr} allows us to map variables to
$\V$-values, so as to put newly bound variables in environments; and
\AgdaField{tm} allows us to extract terms from $\V$-values, as required when
we use the environment to handle a free variable.

@@

The idea of simultaneous renaming is that variables replace variables, whereas
with simultaneous substitution, terms replace variables.
This translates to environments for renaming containing $\sqni$-values
(variables), and environments for substitution containing $\vdash$-values
(terms).
